\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage[hypertexnames=false]{hyperref}
\usepackage{xr-hyper}
\externaldocument[CLC-]{catalan-light-cone}
\newcommand{\mainref}[1]{\ref{CLC-#1}}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{float}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}
\geometry{margin=1in}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{remark}{Remark}[section]

\title{Supplement to \emph{The Catalan Light Cone}}
\author{Paul Fernandez}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This companion document is a standalone build of the appendices/proof sketches
that are also included in the main paper (via \verb|\input|). It is
interpretation-neutral and is kept as a convenient separate compilation target.
\end{abstract}

\appendix

\input{catalan-light-cone-supplement-appendices.tex}

\bibliographystyle{plain}
\begin{thebibliography}{99}

	\bibitem{addario-berry13}
	L.~Addario-Berry, L.~Devroye, and S.~Janson.
	\newblock Sub-Gaussian tail bounds for the width and height of conditioned
	Galton--Watson trees.
	\newblock {\em Annals of Probability}, 41(2):1074--1087, 2013.

	\bibitem{ambjorn01}
	J.~Ambj{\o}rn, J.~Jurkiewicz, and R.~Loll.
	\newblock Dynamically triangulating Lorentzian quantum gravity.
	\newblock {\em Nuclear Physics B}, 610:347--382, 2001.

	\bibitem{ambjorn12}
	J.~Ambj{\o}rn, A.~G{\"o}rlich, J.~Jurkiewicz, and R.~Loll.
	\newblock Nonperturbative quantum gravity.
	\newblock {\em Physics Reports}, 519:127--210, 2012.

	\bibitem{barendregt84}
	H.~P. Barendregt.
	\newblock {\em The Lambda Calculus: Its Syntax and Semantics}.
	\newblock North-Holland, 1984.

	\bibitem{bombelli87}
	L.~Bombelli, J.~Lee, D.~Meyer, and R.~D. Sorkin.
	\newblock Space-time as a causal set.
	\newblock {\em Physical Review Letters}, 59(5):521--524, 1987.

	\bibitem{church33}
	A.~Church.
	\newblock A set of postulates for the foundation of logic.
	\newblock {\em Annals of Mathematics}, 34:839--864, 1933.

	\bibitem{McCarthy1960}
	J.~McCarthy.
	\newblock Recursive functions of symbolic expressions and their computation by
	machine, Part~I.
	\newblock {\em Communications of the ACM}, 3(4):184--195, 1960.

	\bibitem{CurryFeys1958}
	H.~B. Curry and R.~Feys.
	\newblock {\em Combinatory Logic, Vol.~I}.
	\newblock North-Holland, 1958.

	\bibitem{feynman-hibbs65}
	R.~P. Feynman and A.~R. Hibbs.
	\newblock {\em Quantum Mechanics and Path Integrals}.
	\newblock McGraw--Hill, 1965. (Dover reprint, 2010).

	\bibitem{janson07}
	S.~Janson.
	\newblock Brownian excursion area, Wright's constants in graph enumeration, and
	other Brownian areas.
	\newblock {\em Probability Surveys}, 4:80--145, 2007.

	\bibitem{kac49}
	M.~Kac.
	\newblock On distributions of certain Wiener functionals.
	\newblock {\em Transactions of the American Mathematical Society},
	65(1):1--13, 1949.

	\bibitem{le-gall05}
	J.-F. Le~Gall.
	\newblock Random trees and applications.
	\newblock {\em Probability Surveys}, 2:245--311, 2005.

	\bibitem{orus14}
	R.~Or\'us.
	\newblock A practical introduction to tensor networks: Matrix product states and
	projected entangled pair states.
	\newblock {\em Annals of Physics}, 349:117--158, 2014.

	\bibitem{pitman1999}
	J.~Pitman.
	\newblock Brownian motion, bridge, excursion, and meander revisited.
	\newblock {\em Electronic Journal of Probability}, 4:1--32, 1999.

	\bibitem{Schoenfinkel1924}
	M.~Sch{\"o}nfinkel.
	\newblock {\"U}ber die Bausteine der mathematischen Logik.
	\newblock {\em Mathematische Annalen}, 92:305--316, 1924.

	\bibitem{rovelli04}
	C.~Rovelli.
	\newblock {\em Quantum Gravity}.
	\newblock Cambridge University Press, 2004.

	\bibitem{stanley-catalan}
	R.~P. Stanley.
	\newblock {\em Catalan Numbers}.
	\newblock Cambridge University Press, 2015.

		\bibitem{takacs1991}
		L.~Tak{\'a}cs.
		\newblock A {B}ernoulli excursion and its various applications.
		\newblock {\em Advances in Applied Probability}, 23(3):557--585, 1991.

	\bibitem{abhy2018scatteringforms}
	N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan.
	\newblock Scattering forms and the positive geometry of kinematics, color and
	the worldsheet.
	\newblock {\em JHEP} \textbf{05} (2018) 096.
	\newblock arXiv:1711.09102.

	\bibitem{banerjee2018stokes}
	P.~Banerjee, A.~Laddha, and P.~Raman.
	\newblock Stokes polytopes: The positive geometry for $\phi^4$ interactions.
	\newblock arXiv:1811.05904, 2018.

		\bibitem{cheneviere2022linear}
		C.~Chenevi\`ere.
		\newblock Linear intervals in the Tamari and the Dyck lattices and in the
		alt-Tamari posets.
		\newblock arXiv:2209.00418v2, 2022.

	\bibitem{herrmann2022positivegeometry}
	E.~Herrmann and J.~Trnka.
	\newblock Positive geometry of scattering amplitudes.
	\newblock arXiv:2203.13018, 2022.

	\end{thebibliography}
\end{document}
